---
title: "R for Improvers (Script 2): Reshaping, cleaning and summarising data frames"
author: Axel Nohturfft (anohturf@sgul.ac.uk)
date: "`r Sys.Date()`"
output:
  html_notebook:
    code_folding: 'no'
    fig_caption: yes
    highlight: tango
    number_sections: no
    theme: sandstone
    toc: yes
    toc_depth: 3
    toc_float: no
---

**_Please follow the instructions in the README file before starting this tutorial._**

## Objectives  

(1) In this part of the workshop we will use two examples to illustrate how to convert data from a wide to a long (tidy) arrangement and how such data can then be summarised in tabular or graphical format.  
(2) Along the way we will use some functions from the 'dplyr' package to manipulate data frames  

    * Changing column names: `dplyr::rename()`  
    * Extract or remove specific columns: `dplyr::select()`  
    * Sort data frames: `dplyr::arrange()`  
    * Modify or add columns: `dplyr::mutate`  
    * Selecting rows based on specific criteria: `dplyr::filter()`
  

(3) Cleaning: we will look at an example of how to remove rows with missing data points using the `tidyr::drop_na()` function.  
(4) We will show how to perform summary calculations on specific groups of rows within a data frame (`dplyr::group_by` and `dplyr::summarise` functions)

## Introduction: the 'tidy data' concept  
A lot of effort during data analysis projects goes into shaping input data into a format that can then be analysed with standard tools, a process known as '[data wrangling](https://en.wikipedia.org/wiki/Data_wrangling)' or 'data munging'.  

For example, the same exam results can be tabulated in the following two formats:  

|           | Question_1 | Question_2 | Question_3 | Question_4 |
|-----------|------------|------------|------------|------------|
| Student_a | A          | C          | B          | E          |
| Student_b | A          | B          | D          | E          |
| Student_c | A          | C          | D          | B          |


| Student | Question | Answer |
|---------|----------|--------|
| a       | 1        | A      |
| a       | 2        | C      |
| a       | 3        | B      |
| a       | 4        | E      |
| b       | 1        | A      |
| b       | 2        | B      |
| b       | 3        | D      |
| b       | 4        | E      |
| c       | 1        | A      |
| c       | 2        | C      |
| c       | 3        | D      |
| c       | 4        | B      |

The first table can be referred to as _**wide format**_, and the second version can be called the _**long format**_.  

_The term **TIDY DATA** is used in reference to tables arranged in such a way that each row represents an individual sample, and each column represents a variable._  

In the first table above the four question columns all contain the same type of data - students' answers to multiple choice questions. Plus, the row and column headers contain variables (student and question IDs); this table is *not* 'tidy'.  

The data would have to be converted into the format of the second (long-format) table to be conveniently analysed.  

A whole collection of R packages known as the [tidyverse](https://www.tidyverse.org/) are built around the tidy data concept ([Wickham, 2013](https://www.jstatsoft.org/index.php/jss/article/view/v059i10/v59i10.pdf)). These packages are very widely used in the R community.  

## Setup  
```{r chunk options, include=FALSE}
# setting rmarkdown options
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

```{r clean up memory}
# clear all variables that might be stored from a previous analysis:
rm(list=ls(all=T))
```

```{r packages, message=FALSE}
# load packages
library(magrittr)
# help(package="magrittr")
library(readr)
# help(package="readr")
library(dplyr)
# help(package="dplyr")
library(tidyr)
# help(package="tidyr")
library(stringr)
# help(package="stringr")
library(ggplot2)
# help(package="ggplot2")
```

## Time curve of phosphate data  
*(Original data kindly provided by Felix Effah)*  

### Step 1: Load the original data.

* If you check the CSV file in Excel, you will notice that the first column does not have a header, so R makes one up ("X1").  
* Also notice that some data are missing at t0.5 and t1. R inserted NA instead.  

```{r load phosphate, message=FALSE, warning=FALSE}
phosphate <- readr::read_csv("data/phosphate_b.csv")
# View(phosphate)
print(phosphate)
```

```{r results='asis'}
head(phosphate) %>% as.data.frame
```


```{r}
p2 <- phosphate %>% 
  dplyr::rename(Patient = "X1") %>% 
  dplyr::rename(Group = "group") %>% 
  tidyr::pivot_longer(-c("Patient", "Group"), names_to = "Time", values_to = "Phosphate")
p2
```


### Step 2: The first column contains just row numbers, so we can remove it:
```{r}
phosphate.2 <- phosphate %>% 
  dplyr::select(-X1)
# View(phosphate.2)
print(phosphate.2)
```

### Step 3: The values in columns t0 to t5 all contain the same type of data - phosphate concentrations. Let's convert the table to a tidy (long) format using the `pivot_longer()` function from the '[tidyr](https://tidyr.tidyverse.org/)' package.
```{r}
phosphate.3 <- phosphate.2 %>% 
  tidyr::pivot_longer(-group, names_to = "Time", values_to = "Phosphate")
# View(phosphate.3)
print(phosphate.3)
```

### Step 3: For easier inspection let's sort the data frame by group and time:  
```{r}
phosphate.4 <- phosphate.3 %>% 
  dplyr::rename(Group = group) %>% 
  dplyr::arrange(Group, Time)
# View(phosphate.4)
print(phosphate.4)
```

### Step 4: Fix the format of the Time column  
The letter 't' in the Time column is redundant and would prevent us from plotting the data on a numerical time axis. We remove the letter 't' and convert the column to numeric format:  
(The 'stringr' package includes a whole host of useful function to manipulate text.)  

```{r}
phosphate.5 <- phosphate.4 %>% 
  dplyr::mutate(Time = stringr::str_remove(Time, "t")) %>% 
  dplyr::mutate(Time = as.numeric(Time))
# View(phosphate.5)
print(phosphate.5)
```

### Step 5: Cleaning the table  
As noted earlier, the table includes a number of rows with missing data. Let's remove these rows.  

* There are different ways to accomplish this task. One approach using base R would be to use the `complete.cases()` function.  
* Another option would be to use the `dplyr::filter()` in combination with the `is.na()` function.  
* Here we use the most convenient tool: `tidyr::drop_na`.  

```{r}
# drop_na(phosphate.5) would remove rows containing NA in any column
# drop_na(phosphate.5, Phosphate) will remove rows with NA specifically in the Phosphate column
phosphate.6 <- tidyr::drop_na(phosphate.5, Phosphate)
# View(phosphate.6)
print(phosphate.6)
```

*(notice how the new table now has only 259 instead of 264 rows)*  

### Step 6: Draw a boxplot  
Inspection of the data indicates that we have 20 replicates for each combination of group and time point.  
We could illustrate the spread of the data as a boxplot (keeping in mind that boxplots are for categorical x values):
```{r phosphate boxplot}
# To allow ggplot creating a boxplot with time on the x-axis, the Time column has to be
# converted back to a specific text format, called 'factor':
gg1 <- phosphate.6 %>% 
  dplyr::mutate(Time = factor(Time, levels=sort(unique(phosphate.6$Time)))) %>% 
  ggplot(aes(x=Time, y=Phosphate)) + geom_boxplot() + facet_wrap(~Group)
gg1
```

### Step 7: Draw a scatterplot  
Alternatively, one might want to calculate averages for each time point and display the data as a scatterplot.  
In this case we first use the `group_by()` and `summarise()` functions from the tidyverse 'dplyr' package.  
The way this works is that, effectively, dplyr splits the input into a series of smaller data frames as specified in the `group_by()` arguments; the `summarise()` function then calculates statistics on each of the data frame 'chunks' before putting it all together in one output:  

```{r phosphate scatterplot}
phosphate.7 <- phosphate.6 %>% 
  dplyr::group_by(Group, Time) %>% 
  dplyr::summarise(Average = mean(Phosphate),
                   SD = sd(Phosphate)) %>% 
  dplyr::mutate(lower = Average - SD,
                upper = Average + SD)
# View(phosphate.6)
dplyr::select(phosphate.7, -lower, -upper)
```

```{r}
gg2 <- phosphate.7 %>% 
  ggplot(aes(x=Time, y=Average, color=Group)) +
  geom_line() +
  # geom_errorbar(aes(ymin=lower, ymax=upper, width=0.1)) +
  geom_point() +
  labs(x="Time (min)",
       y="Phosphate concentration (mM)")
gg2
```

In "real life" one would have processed the phosphate data in one continuous chain of pipes:
```{r message=FALSE, warning=FALSE}
phosphate.x <- readr::read_csv("data/phosphate_b.csv") %>% 
  dplyr::select(-X1) %>% 
  tidyr::pivot_longer(-group, names_to = "Time", values_to = "Phosphate") %>% 
  dplyr::rename(Group = group) %>% 
  dplyr::arrange(Group, Time) %>% 
  dplyr::mutate(Time = stringr::str_remove(Time, "t")) %>% 
  dplyr::mutate(Time = as.numeric(Time)) %>% 
  dplyr::filter(!is.na(Phosphate))

phosphate.summary <- phosphate.x %>% 
  dplyr::group_by(Group, Time) %>% 
  dplyr::summarise(Average = mean(Phosphate),
                   SD = sd(Phosphate)) %>% 
  dplyr::mutate(lower = Average - SD,
                upper = Average + SD)
```

## EXERCISE  

The file data/SAQ_Exam_Results.csv contains the results of an SAQ exam with 6 questions for 4 students.  
Each question was worth a maximum of 5 points.  

(1) Calculate a percentage mark for each student (Cand.No)  
(2) Calculate the average percent facility for each of the six exam questions (average percentage mark).  

```{r exercise}
a <- readr::read_csv(file="data/SAQ_Exam_Results.csv")
View(a)
b <- tidyr::pivot_longer(a, -Cand.No, names_to = "Question", values_to = "Points")
View(b)
marks <- dplyr::group_by(b, Cand.No) %>% 
  dplyr::summarise(Total_Points = sum(Points)) %>% 
  dplyr::mutate(Mark = (Total_Points / 30)*100) %>% 
  dplyr::mutate(Mark = round(Mark, 1)) %>% 
  dplyr::mutate(Mark = paste0(Mark, "%"))
marks
```

```{r}
facilities <- dplyr::group_by(b, Question) %>% 
  dplyr::summarise(Average_Points = mean(Points)) %>% 
  dplyr::mutate(Facility = (Average_Points / 5)*100) %>% 
  dplyr::mutate(Facility = paste0(Facility, "%"))
facilities
```

## Session info  
```{r}
sessionInfo()
```


