---
title: "R for Improvers (Script 3): Joining (merging) data frames"
author: Axel Nohturfft (anohturf@sgul.ac.uk)
date: "`r Sys.Date()`"
output:
  html_notebook:
    code_folding: 'no'
    fig_caption: yes
    highlight: tango
    number_sections: no
    theme: sandstone
    toc: yes
    toc_depth: 3
    toc_float: no
---

**_Please follow the instructions in the README file before starting this tutorial._**

## Objectives  

(1) This part of the workshop focuses on how to merge (join) data frames  
(2) And we will practise techniques learned earlier (reshaping and manipulating data frames)  

## Setup  
```{r chunk options, include=FALSE}
# setting rmarkdown options
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

```{r clean up memory}
# clear all variables that might be stored from a previous analysis:
rm(list=ls(all=T))
```

```{r packages, message=FALSE}
# load packages
library(magrittr)
# help(package="magrittr")
library(readr)
# help(package="readr")
library(dplyr)
# help(package="dplyr")
library(tidyr)
# help(package="tidyr")
library(stringr)
# help(package="stringr")
library(ggplot2)
# help(package="ggplot2")
```

## Merging / joining data frames  

* A very common task. One example would be to have two tables: one with gene symbols and expression data, and another with the gene symbols and gene names - your goal would be to add the gene names to expression data.  
* In general programming parlance people refer to this manipulation as performing a _**join**_.  
* The function we use in base R is `merge()`.  
* The 'dplyr' package also provides a number of join function that work well with magrittr pipes and other tidyverse functions.  

Here, let's go back to our exam files from the first script...  

### Step 1: Load the blueprint file and select the relevant columns  
```{r open blueprint, message=FALSE}
csv.file <- "data/OSPE_Blueprint.csv"
stopifnot(file.exists(csv.file))
df.blueprint <- readr::read_csv(file=csv.file, col_names = TRUE, comment = "#") %>% 
  dplyr::select(Question, Correct_Answer)
print(df.blueprint)
```

### Step 2: Load the file with the student answers  

* __Note__  
* that the answer data include a question/column (OSPE6) for which there is no match in the blueprint table loaded above.  
* and that the blueprint data included a question (OSPE5) for which there is no match in the results table.  

```{r open tsv, message=FALSE}
tsv.file <- "data/OSPE_Results_Raw.tsv"
stopifnot(file.exists(tsv.file))
df.results <- readr::read_tsv(file=tsv.file, col_names = TRUE, comment = "#")
print(df.results)
```

### Step 3: Reshape student answers into 'tidy' (long) format  
```{r}
results.long <- tidyr::pivot_longer(df.results, -Cand.No, names_to = "Question", values_to = "Answer")
print(results.long)
```

### Step 4: Merge results and blueprint  
```{r}
df.merged <- merge(x=results.long, y=df.blueprint, by="Question", all.x=FALSE, all.y=FALSE) %>% 
  dplyr::select(Cand.No, everything()) %>% 
  dplyr::arrange(Cand.No, desc(Question))
print(df.merged)
```

**Note that the merged table above contains no rows for OSPE5 or OSPE6.** By default, the `merge()` function will only include data for which it finds matches in *both* tables. This is also called an **INNER JOIN**. Decisions as to whether include all rows from one or both input tables is set by the `all.x` and `all.y` arguments to the `merge()` function.  

To illustrate the different options, let's focus on just one of the candidates (1001):  

**INNER JOIN** - Include only matches from both tables:
```{r}
merge(x=results.long, y=df.blueprint, by="Question", all.x=FALSE, all.y=FALSE) %>% 
  dplyr::select(Cand.No, everything()) %>% 
  dplyr::filter(Cand.No == 1001) %>% 
  dplyr::arrange(Question)
```

**LEFT JOIN** - Include all data from table 1 (`all.x=TRUE`), only matching rows from table 2 (`all.y=FALSE`):
```{r}
merge(x=results.long, y=df.blueprint, by="Question", all.x=TRUE, all.y=FALSE) %>% 
  dplyr::select(Cand.No, everything()) %>% 
  dplyr::filter(Cand.No == 1001) %>% 
  dplyr::arrange(Question)
```

**RIGHT JOIN** - Include all data from table 2 (`all.y=TRUE`), only matching rows from table 1 (`all.x=FALSE`):
```{r}
merge(x=results.long, y=df.blueprint, by="Question", all.x=FALSE, all.y=TRUE) %>% 
  dplyr::select(Cand.No, everything()) %>% 
  dplyr::filter(Cand.No %in% c(1001, NA)) %>%
  dplyr::arrange(Question)
```

**OUTER JOIN** - Include all data from table 1 (`all.x=TRUE`), and all data from table 2 (`all.y=TRUE`):
```{r}
merge(x=results.long, y=df.blueprint, by="Question", all.x=TRUE, all.y=TRUE) %>% 
  dplyr::select(Cand.No, everything()) %>% 
  dplyr::filter(Cand.No %in% c(1001, NA)) %>%
  dplyr::arrange(Question)
```

### Step 5: "Mark" the answers  
```{r}
df.marked <- df.merged %>% 
  dplyr::mutate(Answer_Correct = (Answer == Correct_Answer))
print(df.marked)
```


## EXERCISE  

The file `data/Tailleux_TB_MP.csv` contains some microarray-based gene expression data from MTb-infected macrophages for 50 gene probes and 2 time points (0 and 48 hrs). The expression data are log(2) values. (Data from [Tailleux et al., 2008](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0001403))

The file `data/ID_Symbol.csv` lists the corresponding gene symbols for the probe IDs on the microarray chip that was used.

(1) Add the correct gene symbols to the table with gene expression data.  
(2) Calculate the fold change of expression, comparing 48 versus 0 hours.  
(3) Calculate the maximum fold change for each gene (symbol)  

### Step 1: load expression data
```{r expression data, message=FALSE}
# solution here
```

### Step 2: load symbol key
```{r symbol key, message=FALSE}
# solution here
```

### Step 3: merge data frames on 'ID' columns
```{r merge, message=FALSE}
# solution here
```

### Step 4: Calculate fold change
```{r fold change, message=FALSE}
# solution here
```

### Aggregate by gene symbol
```{r}
# solution here
```

## Session info  
```{r}
sessionInfo()
```


